COATi: statistical pairwise alignment of protein coding sequences

### 2024-04-16 supplementary_materials/supplementary_materials.Rmd
Add p-dist to supplemental material

```{r}

summary_table = function(results_file, coati_model) {
    # read results_summary data
    results = read.csv(results_file)

    # extract dseq values for each aligner
    coati = results |> filter(aligner == coati_model) |> select(dseq)
    clustal = results |> filter(aligner == "clustalo") |> select(dseq)
    macse = results |> filter(aligner == "macse") |> select(dseq)
    mafft = results |> filter(aligner == "mafft") |> select(dseq)
    prank = results |> filter(aligner == "prank") |> select(dseq)

    stats_table = data.frame(dseq = c(mean(coati$dseq, na.rm = TRUE),
                          mean(mafft$dseq, na.rm = TRUE),
                          mean(prank$dseq, na.rm = TRUE),
                          mean(macse$dseq, na.rm = TRUE),
                          mean(clustal$dseq, na.rm = TRUE)), row.names = c(coati_model, "mafft", "prank", "macse", "clustalo"))

    ###########################################################################
    # perfect, best, and imperfect alignments
    source("scripts/number_alignments.R")
    alns = num_alns(results_file)

    # fix row and column names and add to stats table
    colnames(alns) = c("Perfect alns", "Best alns", "Imperfect alns")
    alns = alns[c(coati_model, "mafft", "prank", "macse", "clustalo"), ]
    stats_table = cbind(stats_table, alns)

    ############################################################################
    # selection (ka ks)
    source("scripts/kaks.R")

    aligners = c(coati_model, "mafft", "prank", "macse", "clustalo")
    selection = data.frame("F1.pos" = double(),
                           "F1.neg" = double())

    # calculate F1 score for positive and negative selection for each aligner
    for(model in aligners) {
        model_rows = results[which(results$aligner == model), ]
        pos = ps_accuracy(model_rows$ref_omega, model_rows$aln_omega)
        neg = ns_accuracy(model_rows$ref_omega, model_rows$aln_omega)

        selection[nrow(selection) + 1, ] = c(pos, neg)
    }

    # fix row and column names and add to stats table
    colnames(selection) = c("F1 pos selection", "F1 neg selection")
    rownames(selection) = c("coati", "mafft", "prank", "macse", "clustalo")
    stats_table = cbind(stats_table, selection)
    if(coati_model == "tri-mg") {
        coati_model = "Triplet-MG94"
    } else if(coati_model == "tri-ecm") {
        coati_model = "Triplet-ECM"
    } else if(coati_model == "mar-mg") {
        coati_model = "Marginal-MG94"
    } else if(coati_model == "mar-ecm") {
        coati_model = "Marginal-ECM"
    }
    rownames(stats_table) = c(coati_model, "MAFFT", "PRANK*", "MACSE", "ClustalOmega")

    # display stats table
    stats_table
}

wilcoxon_test = function(results_file, coati_model) {
    # read results_summary data
    results = read.csv(results_file)

    # extract dseq values for each aligner
    coati = results$dseq[which(results$aligner == coati_model)]
    clustalomega = results$dseq[which(results$aligner == "clustalo")]
    macse = results$dseq[which(results$aligner == "macse")]
    mafft = results$dseq[which(results$aligner == "mafft")]
    prank = results$dseq[which(results$aligner == "prank")]

    # one-side paired wilcoxon rank test
    wtp = wilcox.test(x = coati, y = prank, paired = TRUE, alternative = "less")
    wtmf = wilcox.test(x = coati, y = mafft, paired = TRUE, alternative = "less")
    wtc = wilcox.test(x = coati, y = clustalomega, paired = TRUE, alternative = "less")
    wtmc = wilcox.test(x = coati, y = macse, paired = TRUE, alternative = "less")
    w = c(wtp$p.value, wtmf$p.value, wtc$p.value, wtmc$p.value)
    w
}
```

@@

We quantified the lengths of gaps produced by each method across the empirical dataset of human-gorilla protein-coding-sequence pairs (Tab. \@ref(tab:gap-table-1)). Here the gap type is either "D" for gaps introduced into the gorilla sequence or "I" for gaps introduced in the human sequence. "COATi" refers to the full COATi FST model with a MG substitution model (i.e. tri-mg). "COATi-rev" refers to the same model, but using gorilla as the reference and human as the non-reference sequence. Gap lengths of 1--6 nucleotides were binned into their respective columns. Gap lengths longer than 6 nucleotides were binned into columns 7+, 8+, and 9+ depending on whether their lengths were 1, 2, or 3 nucleotides longer than a multiple of three.

@@

gg

@@

k <- kable(tab_data1, digits = 4, align = "ccccc", caption="Average K2P distances of alignments for each method.", booktabs = TRUE, linesep = c("\\addlinespace"), format.args = list(nsmall = 4))

k |> kable_styling(latex_options = "HOLD_position")

@@

k <- kable(tab_data2, digits = 4, align = "ccccc", caption="Average K2P distances of alignments for each COATi model.", booktabs = TRUE, linesep = c("\\addlinespace"), format.args = list(nsmall = 4))

>>>>

@@

We quantified the lengths of gaps produced by each method across the empirical
dataset of human-gorilla protein-coding-sequence pairs (Tab. \@ref
(tab:gap-table-1)). Here the gap type is either "D" for gaps introduced into
the gorilla sequence or "I" for gaps introduced in the human sequence. "COATi"
refers to the full COATi FST model with a MG substitution model
(i.e. tri-mg). "COATi-rev" refers to the same model, but using gorilla as the
reference and human as the non-reference sequence. Gap lengths of 1--6
nucleotides were binned into their respective columns. Gap lengths longer than
6 nucleotides were binned into columns 7+, 8+, and 9+ depending on whether
their lengths were 1, 2, or 3 nucleotides longer than a multiple of three.

@@

## Sequence Distances

We quantified the raw sequence distances (p-distance) inferred by different
aligners and COATi models (Fig. \@ref(fig:p-figure-1),
Tab. \@ref(tab:p-table-1), and Tab. \@ref(tab:p-table-2)).

```{r p-figure-1, fig.cap="COATi produced shorter sequence distances than other aligners. Each panel is a scatter plot where the x coordinate is the p-distance from a COATi alignment and y coordinate is the p-distance from the corresponding alignment from another aligner.", warning = FALSE}
p_data <- ALN_DATA |>
    filter(method %in% c(c("clustalo", "coati-tri-mg",
    "macse", "mafft", "prank", "rev-coati-tri-mg"))) |>
    pivot_wider(id_cols=gene, names_from=method, values_from=pdist) |>
    rename(coati = `coati-tri-mg`, `coati-rev` = `rev-coati-tri-mg`) |>
    relocate(gene, coati)

p_data1 <- p_data |> pivot_longer(cols = clustalo:`coati-rev`,
    names_to="method", values_to="other") |>
    mutate(method = fct_relevel(method, "clustalo", "macse", "mafft",
        "prank", "coati-rev")) |>
    mutate(method = fct_recode(method,
        "vs ClustalΩ" = "clustalo",
        "vs MACSE" = "macse",
        "vs MAFFT" = "mafft",
        "vs PRANK" = "prank",
        "vs COATi-rev" = "coati-rev"
    ))

gg <- ggplot(p_data1, aes(x = coati, y = other, color = method)) + facet_wrap(vars(method))
gg <- gg + geom_point(alpha = 0.2, show.legend = FALSE)
gg <- gg + xlab("COATi") + ylab("Other")
gg <- gg + scale_color_manual(values = pal)
gg <- gg + scale_x_continuous(trans = scales::transform_pseudo_log(0.0004, base=10),
    breaks=c(0, 0.001, 0.01, 0.1, 1))
gg <- gg + scale_y_continuous(trans = scales::transform_pseudo_log(0.0004, base=10),
    breaks=c(0, 0.001, 0.01, 0.1, 1))

```{r p-table-1, results = 'asis', message = FALSE}
tab <- ALN_DATA |> summarize(m = mean(pdist, na.rm = TRUE), .by = "method") |>
    rename(Method = method)

k <- kable(tab_data1, digits = 4, align = "ccccc",
  caption="Average p-distance of alignments for each method.",
  booktabs = TRUE, linesep = c("\\addlinespace"),
  format.args = list(nsmall = 4))

```{r p-table-2, results = 'asis', message = FALSE}
tab_data2 <- tab |> filter(Method %in% c("coati-tri-mg",
    "coati-tri-ecm", "coati-mar-mg", "coati-mar-ecm", "coati-dna")) |>
    mutate(Method = fct_relevel(Method, c("coati-tri-mg",
        "coati-tri-ecm", "coati-mar-mg", "coati-mar-ecm", "coati-dna"))) |>
    arrange(Method) |>
    rename(Model = Method) |>
    mutate(Model = fct_recode(Model,
        "TRI-MG" = "coati-tri-mg",
        "TRI-ECM" = "coati-tri-ecm",
        "MAR-MG" = "coati-mar-mg",
        "MAR-ECM" = "coati-mar-ecm",
        "DNA" = "coati-dna"
    ))  |>
    pivot_wider(names_from = "Model", values_from = "m")

k <- kable(tab_data2, digits = 4, align = "ccccc",
    caption="Average p-distance of alignments for each COATi model.",
    booktabs = TRUE, linesep = c("\\addlinespace"),
    format.args = list(nsmall = 4))

@@

k <- kable(tab_data1, digits = 4, align = "ccccc",
  caption="Average K2P-distance of alignments for each method.",
  booktabs = TRUE, linesep = c("\\addlinespace"),
  format.args = list(nsmall = 4))

@@

k <- kable(tab_data2, digits = 4, align = "ccccc",
    caption="Average K2P-distance of alignments for each COATi model.",
    booktabs = TRUE, linesep = c("\\addlinespace"),
    format.args = list(nsmall = 4))

### 2024-04-16 supplementary_materials/supplementary_materials.tex
Revision 1 - Release Candidate 1
Add p-dist to supplemental material

{
\setcounter{tocdepth}{2}
\tableofcontents
}
\hypertarget{aligner-commands}{%
\section{Aligner Commands}\label{aligner-commands}}

@@

\begin{itemize}
\tightlist
\item
  COATi: \texttt{coati\ alignpair\ -m\ tri-mg\ ...}; We used COATi two different ways. The primary way used human as the reference sequence. The secondary way, COATi-rev, used gorilla as the reference, via a wrapper script that reversed the order of sequences before alignment and reversed the order back afterwards.
\item
  ClustalΩ v1.2.4: \texttt{clustalo\ -\/-seqtype=Protein\ ...}; We wrapped ClustalΩ with a script that translates DNA sequences into amino acid sequences (including stops) before alignment. Any codons that were partial or containing ambiguous characters were translated as ``X''. The script then aligned these translated sequences with ClustalΩ, and then created a DNA alignment that was consistent with the amino-acid alignment.
\item
  MACSE v2.06: \texttt{java\ -jar\ macse.jar\ -prog\ alignSequences\ -seq\ human.fasta\ -seq\_lr\ gorilla.fasta\ ...}; We wrapped MACSE with a script that created two temporary fasta files, one containing the human sequence and another containing the gorilla sequence. The human sequence was specified as the reliable sequence and the gorilla sequence was specified as the less-reliable sequence. MACSE uses ``!'' to mark gaps that result from frameshifts, and the wrapper script replaced these with ``-''. Additionally, MACSE sometimes produced columns that only contained gaps, and these columns were removed.
\item
  PRANK v.150803: \texttt{prank\ -codon\ ...}
\item
  MAFFT v7.520: \texttt{mafft\ -\/-preservecase\ -\/-globalpair\ -\/-maxiterate\ 1000\ ...}
\end{itemize}

@@

\hypertarget{fst-alignment-example}{%
\section{FST Alignment Example}\label{fst-alignment-example}}

@@

\hypertarget{empirical-results}{%
\section{Empirical Results}\label{empirical-results}}

\hypertarget{gap-patterns}{%
\subsection{Gap Patterns}\label{gap-patterns}}

\hypertarget{lengths}{%
\subsubsection{Lengths}\label{lengths}}

We quantified the lengths of gaps produced by each method across the empirical dataset of human-gorilla protein-coding-sequence pairs (Tab. \ref{tab:gap-table-1}). Here the gap type is either ``D'' for gaps introduced into the gorilla sequence or ``I'' for gaps introduced in the human sequence. ``COATi'' refers to the full COATi FST model with a MG substitution model (i.e.~tri-mg). ``COATi-rev'' refers to the same model, but using gorilla as the reference and human as the non-reference sequence. Gap lengths of 1--6 nucleotides were binned into their respective columns. Gap lengths longer than 6 nucleotides were binned into columns 7+, 8+, and 9+ depending on whether their lengths were 1, 2, or 3 nucleotides longer than a multiple of three.

@@

\begin{table}[H]

\caption{\label{tab:gap-table-1}Number of gaps introduced by each method separated by length and type.}
\centering
\begin{tabular}[t]{lcrrrrrrrrr}
\toprule
\multicolumn{2}{c}{ } & \multicolumn{9}{c}{Gap Lengths} \\
\cmidrule(l{3pt}r{3pt}){3-11}
Method & Gap Type & 1 & 2 & 3 & 4 & 5 & 6 & 7+ & 8+ & 9+\\
\midrule
COATi & D & 106 & 99 & 651 & 94 & 94 & 226 & 1461 & 1386 & 3240\\
COATi & I & 87 & 73 & 525 & 80 & 97 & 239 & 1399 & 1315 & 3105\\
\addlinespace
COATi-rev & D & 105 & 99 & 634 & 89 & 96 & 217 & 1425 & 1348 & 3169\\
COATi-rev & I & 82 & 79 & 513 & 80 & 97 & 231 & 1360 & 1288 & 3031\\
\addlinespace
ClustalΩ & D & 1 & 0 & 887 & 1 & 1 & 400 & 3 & 7 & 3910\\
ClustalΩ & I & 0 & 0 & 857 & 0 & 1 & 424 & 3 & 2 & 3747\\
\addlinespace
MACSE & D & 396 & 306 & 1237 & 5 & 3 & 657 & 71 & 54 & 4784\\
MACSE & I & 60 & 27 & 1133 & 5 & 9 & 666 & 78 & 97 & 4381\\
\addlinespace
MAFFT & D & 204 & 156 & 714 & 147 & 108 & 276 & 544 & 509 & 2680\\
MAFFT & I & 187 & 154 & 708 & 112 & 92 & 295 & 454 & 424 & 2642\\
\addlinespace
PRANK & D & 0 & 0 & 552 & 0 & 0 & 167 & 0 & 0 & 4812\\
PRANK & I & 0 & 0 & 467 & 0 & 0 & 178 & 0 & 0 & 4686\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[H]

\caption{\label{tab:gap-table-2}Number of gaps introduced by different COATi models separated by length and type.}
\centering
\begin{tabular}[t]{lcrrrrrrrrr}
\toprule
\multicolumn{2}{c}{ } & \multicolumn{9}{c}{Gap Lengths} \\
\cmidrule(l{3pt}r{3pt}){3-11}
Model & Gap Type & 1 & 2 & 3 & 4 & 5 & 6 & 7+ & 8+ & 9+\\
\midrule
TRI-MG & D & 106 & 99 & 651 & 94 & 94 & 226 & 1461 & 1386 & 3240\\
TRI-MG & I & 87 & 73 & 525 & 80 & 97 & 239 & 1399 & 1315 & 3105\\
\addlinespace
TRI-ECM & D & 93 & 96 & 665 & 125 & 122 & 257 & 1438 & 1441 & 3203\\
TRI-ECM & I & 86 & 103 & 540 & 105 & 122 & 254 & 1358 & 1370 & 3099\\
\addlinespace
MAR-MG & D & 102 & 106 & 646 & 95 & 89 & 224 & 1465 & 1399 & 3226\\
MAR-MG & I & 84 & 83 & 526 & 80 & 101 & 238 & 1406 & 1313 & 3087\\
\addlinespace
MAR-ECM & D & 110 & 110 & 653 & 107 & 89 & 224 & 1424 & 1389 & 3250\\
MAR-ECM & I & 86 & 89 & 523 & 77 & 99 & 249 & 1408 & 1310 & 3128\\
\addlinespace
DNA & D & 101 & 103 & 646 & 95 & 91 & 224 & 1446 & 1409 & 3206\\
DNA & I & 79 & 79 & 520 & 78 & 99 & 239 & 1369 & 1316 & 3107\\
\bottomrule
\end{tabular}
\end{table}

\hypertarget{phases}{%
\subsubsection{Phases}\label{phases}}

@@

\begin{table}[H]

\caption{\label{tab:gap-table-3}Number of gaps introduced by each alignment method separated by phase.}
\centering
\begin{tabular}[t]{lrrr}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{3}{c}{Gap Phases} \\
\cmidrule(l{3pt}r{3pt}){2-4}
Method & 1 & 2 & 3\\
\midrule
COATi & 4493 & 3962 & 5822\\
\addlinespace
COATi-rev & 3992 & 3775 & 6176\\
\addlinespace
ClustalΩ & 8 & 6 & 10230\\
\addlinespace
MACSE & 455 & 497 & 13017\\
\addlinespace
MAFFT & 2317 & 2631 & 5458\\
\addlinespace
PRANK & 0 & 0 & 10862\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[H]

\caption{\label{tab:gap-table-4}Number of gaps introduced by different COATi models separated by phase.}
\centering
\begin{tabular}[t]{lrrr}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{3}{c}{Gap Phases} \\
\cmidrule(l{3pt}r{3pt}){2-4}
Model & 1 & 2 & 3\\
\midrule
TRI-MG & 4493 & 3962 & 5822\\
\addlinespace
TRI-ECM & 4160 & 4678 & 5639\\
\addlinespace
MAR-MG & 4130 & 3961 & 6179\\
\addlinespace
MAR-ECM & 4071 & 3863 & 6391\\
\addlinespace
DNA & 4228 & 3974 & 6005\\
\bottomrule
\end{tabular}
\end{table}

\hypertarget{homology-patterns}{%
\subsection{Homology Patterns}\label{homology-patterns}}

@@

\begin{table}[H]

\caption{\label{tab:gap-table-5}Average homology percentages of alignments separated by alignment method.}
\centering
\begin{tabular}[t]{rccc}
\toprule
Method & Matches & Mismatches & Gaps\\
\midrule
COATi & 95.93\% & 0.79\% & 3.28\%\\
\addlinespace
COATi-rev & 95.95\% & 0.79\% & 3.25\%\\
\addlinespace
ClustalΩ & 95.93\% & 1.54\% & 2.52\%\\
\addlinespace
MACSE & 96.13\% & 1.33\% & 2.54\%\\
\addlinespace
MAFFT & 96.12\% & 1.36\% & 2.52\%\\
\addlinespace
PRANK & 95.64\% & 0.84\% & 3.52\%\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[H]

\caption{\label{tab:gap-table-6}Average homology percentages of alignments separated by COATi model.}
\centering
\begin{tabular}[t]{rccc}
\toprule
Model & Matches & Mismatches & Gaps\\
\midrule
TRI-MG & 95.93\% & 0.79\% & 3.28\%\\
\addlinespace
TRI-ECM & 95.90\% & 0.79\% & 3.30\%\\
\addlinespace
MAR-MG & 95.93\% & 0.79\% & 3.28\%\\
\addlinespace
MAR-ECM & 95.94\% & 0.80\% & 3.26\%\\
\addlinespace
DNA & 95.93\% & 0.79\% & 3.27\%\\
\bottomrule
\end{tabular}
\end{table}

@@

\hypertarget{evolutionary-distances}{%
\subsection{Evolutionary Distances}\label{evolutionary-distances}}

@@

\begin{table}[H]

\caption{\label{tab:k2p-table-1}Average K2P distances of alignments for each method.}
\centering
\begin{tabular}[t]{cccccc}
\toprule
COATi & COATi-rev & ClustalΩ & MACSE & MAFFT & PRANK\\
\midrule
0.0084 & 0.0084 & 0.0178 & 0.0149 & 0.0154 & 0.0092\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[H]

\caption{\label{tab:k2p-table-2}Average K2P distances of alignments for each COATi model.}
\centering
\begin{tabular}[t]{ccccc}
\toprule
TRI-MG & TRI-ECM & MAR-MG & MAR-ECM & DNA\\
\midrule
0.0084 & 0.0084 & 0.0084 & 0.0084 & 0.0084\\
\bottomrule
\end{tabular}
\end{table}

\hypertarget{benchmark-results}{%
\section{Benchmark Results}\label{benchmark-results}}

\hypertarget{alignment-distances}{%
\subsection{Alignment Distances}\label{alignment-distances}}

@@

\begin{figure}
\centering
\includegraphics{supplementary_materials_files/figure-latex/bench-pcoa-all-1.pdf}
\caption{\label{fig:bench-pcoa-all}Different COATi produce accurate alignments that are also similar to one another.}
\end{figure}

>>>>

{
\setcounter{tocdepth}{2}
\tableofcontents
}
\section{Aligner Commands}\label{aligner-commands}

@@

\begin{itemize}
\tightlist
\item
  COATi: \texttt{coati\ alignpair\ -m\ tri-mg\ ...}; We used COATi two different ways. The
  primary way used human as the reference sequence. The secondary way,
  COATi-rev, used gorilla as the reference, via a wrapper script that reversed
  the order of sequences before alignment and reversed the order back
  afterwards.
\item
  ClustalΩ v1.2.4: \texttt{clustalo\ -\/-seqtype=Protein\ ...}; We wrapped ClustalΩ with a
  script that translates DNA sequences into amino acid sequences (including
  stops) before alignment. Any codons that were partial or containing ambiguous
  characters were translated as ``X''. The script then aligned these translated
  sequences with ClustalΩ, and then created a DNA alignment that was consistent
  with the amino-acid alignment.
\item
  MACSE v2.06: \texttt{java\ -jar\ macse.jar\ -prog\ alignSequences\ -seq\ human.fasta\ \ \ -seq\_lr\ gorilla.fasta\ ...}; We wrapped MACSE with a script that created two
  temporary fasta files, one containing the human sequence and another
  containing the gorilla sequence. The human sequence was specified as the
  reliable sequence and the gorilla sequence was specified as the less-reliable
  sequence. MACSE uses ``!'' to mark gaps that result from frameshifts, and the
  wrapper script replaced these with ``-''. Additionally, MACSE sometimes produced
  columns that only contained gaps, and these columns were removed.
\item
  PRANK v.150803: \texttt{prank\ -codon\ ...}
\item
  MAFFT v7.520: \texttt{mafft\ -\/-preservecase\ -\/-globalpair\ -\/-maxiterate\ 1000\ ...}
\end{itemize}

@@

\section{FST Alignment Example}\label{fst-alignment-example}

@@

\section{Empirical Results}\label{empirical-results}

\subsection{Gap Patterns}\label{gap-patterns}

\subsubsection{Lengths}\label{lengths}

We quantified the lengths of gaps produced by each method across the empirical
dataset of human-gorilla protein-coding-sequence pairs (Tab. @ref
(tab:gap-table-1)). Here the gap type is either ``D'' for gaps introduced into
the gorilla sequence or ``I'' for gaps introduced in the human sequence. ``COATi''
refers to the full COATi FST model with a MG substitution model
(i.e.~tri-mg). ``COATi-rev'' refers to the same model, but using gorilla as the
reference and human as the non-reference sequence. Gap lengths of 1--6
nucleotides were binned into their respective columns. Gap lengths longer than
6 nucleotides were binned into columns 7+, 8+, and 9+ depending on whether
their lengths were 1, 2, or 3 nucleotides longer than a multiple of three.

@@

\begin{table}[H]
\centering
\caption{\label{tab:gap-table-1}Number of gaps introduced by each method separated by length and type.}
\centering
\begin{tabular}[t]{lcrrrrrrrrr}
\toprule
\multicolumn{2}{c}{ } & \multicolumn{9}{c}{Gap Lengths} \\
\cmidrule(l{3pt}r{3pt}){3-11}
Method & Gap Type & 1 & 2 & 3 & 4 & 5 & 6 & 7+ & 8+ & 9+\\
\midrule
COATi & D & 106 & 99 & 651 & 94 & 94 & 226 & 1461 & 1386 & 3240\\
COATi & I & 87 & 73 & 525 & 80 & 97 & 239 & 1399 & 1315 & 3105\\
\addlinespace
COATi-rev & D & 105 & 99 & 634 & 89 & 96 & 217 & 1425 & 1348 & 3169\\
COATi-rev & I & 82 & 79 & 513 & 80 & 97 & 231 & 1360 & 1288 & 3031\\
\addlinespace
ClustalΩ & D & 1 & 0 & 887 & 1 & 1 & 400 & 3 & 7 & 3910\\
ClustalΩ & I & 0 & 0 & 857 & 0 & 1 & 424 & 3 & 2 & 3747\\
\addlinespace
MACSE & D & 396 & 306 & 1237 & 5 & 3 & 657 & 71 & 54 & 4784\\
MACSE & I & 60 & 27 & 1133 & 5 & 9 & 666 & 78 & 97 & 4381\\
\addlinespace
MAFFT & D & 204 & 156 & 714 & 147 & 108 & 276 & 544 & 509 & 2680\\
MAFFT & I & 187 & 154 & 708 & 112 & 92 & 295 & 454 & 424 & 2642\\
\addlinespace
PRANK & D & 0 & 0 & 552 & 0 & 0 & 167 & 0 & 0 & 4812\\
PRANK & I & 0 & 0 & 467 & 0 & 0 & 178 & 0 & 0 & 4686\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[H]
\centering
\caption{\label{tab:gap-table-2}Number of gaps introduced by different COATi models separated by length and type.}
\centering
\begin{tabular}[t]{lcrrrrrrrrr}
\toprule
\multicolumn{2}{c}{ } & \multicolumn{9}{c}{Gap Lengths} \\
\cmidrule(l{3pt}r{3pt}){3-11}
Model & Gap Type & 1 & 2 & 3 & 4 & 5 & 6 & 7+ & 8+ & 9+\\
\midrule
TRI-MG & D & 106 & 99 & 651 & 94 & 94 & 226 & 1461 & 1386 & 3240\\
TRI-MG & I & 87 & 73 & 525 & 80 & 97 & 239 & 1399 & 1315 & 3105\\
\addlinespace
TRI-ECM & D & 93 & 96 & 665 & 125 & 122 & 257 & 1438 & 1441 & 3203\\
TRI-ECM & I & 86 & 103 & 540 & 105 & 122 & 254 & 1358 & 1370 & 3099\\
\addlinespace
MAR-MG & D & 102 & 106 & 646 & 95 & 89 & 224 & 1465 & 1399 & 3226\\
MAR-MG & I & 84 & 83 & 526 & 80 & 101 & 238 & 1406 & 1313 & 3087\\
\addlinespace
MAR-ECM & D & 110 & 110 & 653 & 107 & 89 & 224 & 1424 & 1389 & 3250\\
MAR-ECM & I & 86 & 89 & 523 & 77 & 99 & 249 & 1408 & 1310 & 3128\\
\addlinespace
DNA & D & 101 & 103 & 646 & 95 & 91 & 224 & 1446 & 1409 & 3206\\
DNA & I & 79 & 79 & 520 & 78 & 99 & 239 & 1369 & 1316 & 3107\\
\bottomrule
\end{tabular}
\end{table}

\subsubsection{Phases}\label{phases}

@@

\begin{table}[H]
\centering
\caption{\label{tab:gap-table-3}Number of gaps introduced by each alignment method separated by phase.}
\centering
\begin{tabular}[t]{lrrr}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{3}{c}{Gap Phases} \\
\cmidrule(l{3pt}r{3pt}){2-4}
Method & 1 & 2 & 3\\
\midrule
COATi & 4493 & 3962 & 5822\\
\addlinespace
COATi-rev & 3992 & 3775 & 6176\\
\addlinespace
ClustalΩ & 8 & 6 & 10230\\
\addlinespace
MACSE & 455 & 497 & 13017\\
\addlinespace
MAFFT & 2317 & 2631 & 5458\\
\addlinespace
PRANK & 0 & 0 & 10862\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[H]
\centering
\caption{\label{tab:gap-table-4}Number of gaps introduced by different COATi models separated by phase.}
\centering
\begin{tabular}[t]{lrrr}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{3}{c}{Gap Phases} \\
\cmidrule(l{3pt}r{3pt}){2-4}
Model & 1 & 2 & 3\\
\midrule
TRI-MG & 4493 & 3962 & 5822\\
\addlinespace
TRI-ECM & 4160 & 4678 & 5639\\
\addlinespace
MAR-MG & 4130 & 3961 & 6179\\
\addlinespace
MAR-ECM & 4071 & 3863 & 6391\\
\addlinespace
DNA & 4228 & 3974 & 6005\\
\bottomrule
\end{tabular}
\end{table}

\subsection{Homology Patterns}\label{homology-patterns}

@@

\begin{table}[H]
\centering
\caption{\label{tab:gap-table-5}Average homology percentages of alignments separated by alignment method.}
\centering
\begin{tabular}[t]{rccc}
\toprule
Method & Matches & Mismatches & Gaps\\
\midrule
COATi & 95.93\% & 0.79\% & 3.28\%\\
\addlinespace
COATi-rev & 95.95\% & 0.79\% & 3.25\%\\
\addlinespace
ClustalΩ & 95.93\% & 1.54\% & 2.52\%\\
\addlinespace
MACSE & 96.13\% & 1.33\% & 2.54\%\\
\addlinespace
MAFFT & 96.12\% & 1.36\% & 2.52\%\\
\addlinespace
PRANK & 95.64\% & 0.84\% & 3.52\%\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[H]
\centering
\caption{\label{tab:gap-table-6}Average homology percentages of alignments separated by COATi model.}
\centering
\begin{tabular}[t]{rccc}
\toprule
Model & Matches & Mismatches & Gaps\\
\midrule
TRI-MG & 95.93\% & 0.79\% & 3.28\%\\
\addlinespace
TRI-ECM & 95.90\% & 0.79\% & 3.30\%\\
\addlinespace
MAR-MG & 95.93\% & 0.79\% & 3.28\%\\
\addlinespace
MAR-ECM & 95.94\% & 0.80\% & 3.26\%\\
\addlinespace
DNA & 95.93\% & 0.79\% & 3.27\%\\
\bottomrule
\end{tabular}
\end{table}

@@

\subsection{Sequence Distances}\label{sequence-distances}

We quantified the raw sequence distances (p-distance) inferred by different
aligners and COATi models (Fig. \ref{fig:p-figure-1},
Tab. \ref{tab:p-table-1}, and Tab. \ref{tab:p-table-2}).

\begin{figure}
\centering
\includegraphics{supplementary_materials_files/figure-latex/p-figure-1-1.pdf}
\caption{\label{fig:p-figure-1}COATi produced shorter sequence distances than other aligners. Each panel is a scatter plot where the x coordinate is the p-distance from a COATi alignment and y coordinate is the p-distance from the corresponding alignment from another aligner.}
\end{figure}

\begin{table}[H]
\centering
\caption{\label{tab:p-table-1}Average p-distance of alignments for each method.}
\centering
\begin{tabular}[t]{cccccc}
\toprule
COATi & COATi-rev & ClustalΩ & MACSE & MAFFT & PRANK\\
\midrule
0.0083 & 0.0083 & 0.0168 & 0.0142 & 0.0147 & 0.0089\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[H]
\centering
\caption{\label{tab:p-table-2}Average p-distance of alignments for each COATi model.}
\centering
\begin{tabular}[t]{ccccc}
\toprule
TRI-MG & TRI-ECM & MAR-MG & MAR-ECM & DNA\\
\midrule
0.0083 & 0.0083 & 0.0083 & 0.0083 & 0.0083\\
\bottomrule
\end{tabular}
\end{table}

\subsection{Evolutionary Distances}\label{evolutionary-distances}

@@

\begin{table}[H]
\centering
\caption{\label{tab:k2p-table-1}Average K2P-distance of alignments for each method.}
\centering
\begin{tabular}[t]{cccccc}
\toprule
COATi & COATi-rev & ClustalΩ & MACSE & MAFFT & PRANK\\
\midrule
0.0084 & 0.0084 & 0.0178 & 0.0149 & 0.0154 & 0.0092\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[H]
\centering
\caption{\label{tab:k2p-table-2}Average K2P-distance of alignments for each COATi model.}
\centering
\begin{tabular}[t]{ccccc}
\toprule
TRI-MG & TRI-ECM & MAR-MG & MAR-ECM & DNA\\
\midrule
0.0084 & 0.0084 & 0.0084 & 0.0084 & 0.0084\\
\bottomrule
\end{tabular}
\end{table}

\section{Benchmark Results}\label{benchmark-results}

\subsection{Alignment Distances}\label{alignment-distances}

@@

\begin{figure}
\centering
\includegraphics{supplementary_materials_files/figure-latex/bench-pcoa-all-1.pdf}
\caption{\label{fig:bench-pcoa-all}Different COATi models produce accurate alignments that are also similar to one another.}
\end{figure}
